---
title: "Nonparametric methods for Survival Analysis"
---

# One sample

```{r}
library(survival)
```

## Entering right-censored data in R
```{r}
dat <- data.frame(ratID = paste0("rat", 1:5),
                  time = c(55, 50, 70, 120, 110),
                  status = c(0, 1, 1, 0, 1))
```

## Kaplan-Meyer estimator
```{r}
fit.KM <- survfit(Surv(time, status) ~ 1, data = dat)
summary(fit.KM)
```

```{r}
plot(fit.KM, mark.time = TRUE,
     main = "Kaplan-Meier estimator",
     ylab = "Survival probability",
     xlab = "time (seconds)")
```

Question: what is the median survival time?
```{r}
fit.KM
```

## Nelson-AAlen estimator
```{r}
fit.NA <- survfit(Surv(time, status) ~ 1, data = dat, type = "fh")
summary(fit.NA)
```

```{r}
fit.NA
```

## Case study: the Xelox trial

```{r}
library(asaur)
dat <- gastricXelox
```

How many events, how many censored data points?
```{r}
table(dat$delta)
```

How the Progress Free Survival times data looks like (ignoring censoring info)?
```{r}
hist(dat$timeWeeks * 7 / 365.25)
```

### Kaplan-Meyer estimator
```{r}
fit.KM <- survfit(Surv(timeWeeks, delta) ~ 1, data = dat)
summary(fit.KM)
```

```{r}
plot(fit.KM)
```

Time in weeks might be cumbersome to read: we can re-express it in years
```{r}
dat$timeYears <- dat$timeWeeks * 7 / 365.25
fit.KM <- survfit(Surv(timeYears, delta) ~ 1, data = dat, conf.type = "log-log")
summary(fit.KM)
plot(fit.KM)
```

### Median survival
Question: what is the median survival time?
```{r}
fit.KM
```

Note that the definition of censoring depends on what's the quantity of interest.
If we're interested in measuring the follow-up time, delta is to be 'inverted':
```{r}
dat$delta_followUp <- 1 - dat$delta
fit.followUp <- survfit(Surv(timeYears, delta_followUp) ~ 1, data = dat, conf.type = "log-log")
fit.followUp
```

# Nonparametric comparison of two samples

## Entering right-censored data in R
```{r}
dat <- data.frame(ratID = paste0("rat", 1:5),
                  time = c(55, 50, 70, 120, 110),
                  status = c(0, 1, 1, 0, 1),
                  group = c(0, 1, 0, 1, 1))
```

## The logrank test
```{r}
fit.logrank <- survdiff(Surv(time, status) ~ group, data = dat)
fit.logrank
```

## Case study: the pancreatic dataset

```{r}
library(asaur)

dat <- pancreatic
head(dat)
```

* M: metastatic
* LA: locally advanced

This dataset requires some preprocessing before proper survival analysis.

1. parse 'onstudy', 'progression' and 'death' dates correctly
2. compute progression free survival times and overall survival times (this dataset has no censored data)

### step 1: parse dates
Check the manual page of 'as.Date'
```{r}
fmt <- "%m/%d/%Y"
dat <- within(dat, {
  onstudy <- as.Date(as.character(onstudy), format = fmt)
  progression <- as.Date(as.character(progression), format = fmt)
  death <- as.Date(as.character(death), format = fmt)
})
head(dat)
```

### step 2: compute survival times
```{r}
dat <- within(dat, {
  OS <- difftime(death, onstudy, units = "days")
  PFS <- ifelse(!is.na(progression), difftime(progression, onstudy, units = "days"), OS)
})
```
Note: OS and PFS are expressed in days. We want them in months:
```{r}
dat$OS <- as.numeric(dat$OS) / 30.5
dat$PFS <- as.numeric(dat$PFS) / 30.5
```

### compare PFS in the 2 disease groups

As we have no censoring, we can produce use simple boxplots:
```{r}
library(ggplot2)
```

```{r}
ggplot(dat, aes(stage, PFS)) +
  geom_boxplot() +
  theme_bw()
```

more generally, Kaplan-Meier estimates:
```{r}
fit.KM <- survfit(Surv(PFS) ~ stage, data = dat, conf.type = "log-log")
plot(fit.KM, col = 1:2)
```
Red is a worst survival than black. Became the same at 10 months. The metastatics have a worste time in the first 3 months.

```{r}
fit.KM
```

### The logrank test
```{r}
survdiff(Surv(PFS) ~ stage, data = dat)
```

What's the estimated probability of not experiencing a cancer progression for (at least) 1 year?
```{r}
summary(fit.KM, time = 12)
```

It is similar in the 2 groups, namely between 13% and 15%.
Said otherwise, chances are high that the cancer is going to make a comeback
within one year.

Can you repeat the analysis above, this time for OS?

### compare OS in the 2 disease groups

Visualizing:
```{r}
ggplot(dat, aes(stage, OS)) +
  geom_boxplot() +
  theme_bw()
```

more generally, Kaplan-Meier estimates:
```{r}
fit.KM <- survfit(Surv(OS) ~ stage, data = dat, conf.type = "log-log")
plot(fit.KM, col = 1:2)
```

```{r}
fit.KM
```

### The logrank test
```{r}
survdiff(Surv(OS) ~ stage, data = dat)
```

What's the estimated probability of not experiencing a cancer progression for (at least) 1 year?
```{r}
summary(fit.KM, time = 12)
```


## Stratified logrank test: pharmacoSmoking dataset

### The data
```{r}
dat <- pharmacoSmoking
head(dat)
```

```{r}
summary(dat)
```

Question: do the 2 treatment group differ significantly in terms of survival to relapse?
```{r}
survdiff(Surv(ttr, relapse) ~ grp, data = dat)
```
p < alpha : we can reject null hypothesis "the 2 treatment groups do not differ significantly in terms of survival to relapse"
thus they do differ significantly

Critique: the 2 groups have different age distribution, which might confound our results.
Lets investigate:
```{r}
with(dat, prop.table(table(grp, ageGroup2), 1))
```


```{r}
with(dat, mosaicplot(table(grp, ageGroup2)))
```

### stratified logrank test
```{r}
survdiff(Surv(ttr, relapse) ~ grp + strata(ageGroup2), data = dat)
```
p-value indicates there is still statisticall difference even when stratifying by age.

### extra
```{r}
fit.4 <- survfit(Surv(ttr, relapse) ~ grp + employment, data = dat)
fit.4
```

```{r}
plot(fit.4, col = 1:6)
legend("topright", lty = 1, col = 1:6, legend = names(fit.4$strata))
```
We could see an impact of the employment situation. But in the end, combination grp impacts clearly the ability of people not re-start smoking.

The 3 'combination' curves seem all higher than the 3 'patchOnly' curves. Lets make a stratified test:

```{r}
survdiff(Surv(ttr, relapse) ~ grp + strata(employment), data = dat)
```
It looks like working as p-value is small, so even with the employment status we still have a significant impact of combination therapy

## EXERCISE - analyse by employment, binary

Use subset function to generate a combination+employment(pt/ft) analysis by deleting the "employment="other"" level:
```{r}
subdat <- droplevels(subset(dat, employment != "other"))
```

```{r}
table(subdat$employment)
```
We managed to delete "other" flag employment:
  - subset != omit them
  - droplevels = make the level disappear under employment

```{r}
fit.KM <- survfit(Surv(ttr, relapse) ~ grp + employment, data = subdat)
fit.KM
```
With patch therapy, full-time employment is noticeable in resisting to the temptation where with combination the difference fades.

```{r}
survdiff(Surv(ttr, relapse) ~ grp, data = subset(subdat, employment == "pt"))
```
The difference between "combination" & "patchOnly" groups is not statistically different as p-value > alpha


Stratified test:
```{r}
survdiff(Surv(ttr, relapse) ~ grp + strata(employment), data = subdat)
```