---
title: "Cox regression"
---

# A manually worked out, simple example: two groups

## Load libraries
```{r}
library(reshape)
library(maxLik)
library(survival)
```

## Data definition

Lets enter the data in R:
```{r}
dat <- data.frame(ratID = paste0("rat", 1:5),
                  time = c(55, 50, 70, 120, 110),
                  failure = c(0, 1, 1, 0, 1),
                  group = c(0, 1, 0, 1, 1))
```

Total number of failures D:
```{r}
sum(dat$failure)
```

For convenience, rename 'group' to 'x':
```{r}
dat <- rename(dat, c(group = 'x'))
dat
```

We also define an auxiliary data.frame containing events only:
```{r}
dat.events <- subset(dat, failure == 1)
```

## Partial log-likelihood function

Lets define the partial (log-)likelihood function
```{r}
pLogLik <- function(beta) {
  numerator <- with(dat.events, x * beta)
  denominator <- rep(NA_real_, length(numerator))
  for(j in seq_along(denominator)) {
    risk_set <- subset(dat, time >= dat.events[j, "time"])
    theta_j <- with(risk_set, exp(x * beta))
    denominator[j] <- log(sum(theta_j))
  }
  return(sum(numerator - denominator))
}
```

```{r}
debugonce(pLogLik)
pLogLik(2)
```


We can plot it:
```{r}
f <- Vectorize(pLogLik)
curve(f, from = -4, to = 4)
```

## Maximum partial-Likelihood estimation
```{r}
fit.ML <- maxLik(pLogLik, start = c(beta = 0))
summary(fit.ML)
```

With the `coxph` function:
```{r}
fit.cph <- coxph(Surv(time, failure) ~ x, data = dat)
summary(fit.cph)
```

We can reproduce the Likelihood-ratio test:
```{r}
LRT <- 2 * (fit.ML$maximum - pLogLik(0))
data.frame(LRT = LRT,
           pvalue = pchisq(LRT, df = 1, lower.tail = FALSE))
```

The Wald test is already in the `maxLik` summary output.

# A manually worked out, simple example: one continuous covariate

```{r}
dat <- data.frame(time = c(6, 7, 10, 15, 19, 25),
                  event = c(1, 0, 1, 1, 0, 1),
                  age = c(67, 62, 34, 41, 46, 28))
```

```{r}
fit <- coxph(Surv(time, event) ~ age, data = dat)
summary(fit)
```

We might express age in decades:
```{r}
dat$age_dec <- dat$age / 10
summary(coxph(Surv(time, event) ~ age_dec, data = dat))
```

# Case study: the pharmacoSmoking dataset

## Load the data
```{r}
library(asaur)
dat <- pharmacoSmoking
head(dat)
```

## Fit the Cox model
```{r}
fit <- coxph(Surv(ttr, relapse) ~ employment + age + gender + priorAttempts, data = dat)
summary(fit)
```

```{r}
dat$score <- predict(fit, newdata = dat, type = "lp")
```

```{r}
dat <- dat[order(dat$score), ]
head(dat)
tail(dat)
```


```{r}
fit <- coxph(Surv(ttr, relapse) ~ grp + employment, data = dat)
summary(fit)
```


We can change the contrasts as we see fit:
```{r}
dat$grp <- relevel(dat$grp, ref = "patchOnly")
dat$employment <- relevel(dat$employment, ref = "other")
fit <- update(fit)
summary(fit)
```

# Case study: the lung cancer dataset

## Load the data
```{r}
library(survival)

dat <- lung
dat$sex <- factor(dat$sex)
```

```{r}
summary(dat)
```

```{r}
table(dat$status)
```

```{r}
dat$event <- dat$status - 1
with(dat, table(status, event))
```

## removing missing values
```{r}
d1 <- subset(dat, select = c(time, event, age, sex, ph.ecog, ph.karno, pat.karno))
d <- na.omit(d1)
summary(d)
```

## Survival Estimation

```{r}
fit.KM <- survfit(Surv(time, event) ~ 1, data = d)
fit.KM
```

Median survival was 329 days (95 perc. C.I.: 286-364).

## Impact of ph.ecog on Survival

```{r}
table(d$ph.ecog)
```

```{r}
d$ph.ecog_binary <- (d$ph.ecog > 0) * 1
with(d, table(ph.ecog, ph.ecog_binary))
```

```{r}
fit2 <- survfit(Surv(time, event) ~ ph.ecog_binary, data = d)
fit2
```

```{r}
plot(fit2, col = 1:2)
```

```{r}
survdiff(Surv(time, event) ~ ph.ecog_binary, data = d)
```

```{r}
fit.cph <- coxph(Surv(time, event) ~ ph.ecog_binary, data = d)
summary(fit.cph)
```


## Cox regression: predictions
```{r}
fit.cph <- coxph(Surv(time, status) ~ sex, data = dat)

pred.cph <- survfit(fit.cph, newdata = data.frame(sex = factor(1:2)),
                    type = "aalen")

plot(pred.cph, col = 1:2)
```

How does the proportional hazards assumption hold?
```{r}
plot(pred.NA, fun = "cloglog", col = 1:2)
```

```{r}
plot(pred.cph, fun = "cloglog", col = 1:2)
```
