---
title: "Cox Model Building and Diagnostics"
---

# Model building
## Load the data

```{r}
library(survival)
library(asaur)

dat <- pharmacoSmoking
```

## The 4 candidate models
```{r}
M0 <- coxph(Surv(ttr, relapse) ~ 1, data = dat)
MA <- coxph(Surv(ttr, relapse) ~ ageGroup4, data = dat)
MB <- coxph(Surv(ttr, relapse) ~ employment, data = dat)
MC <- coxph(Surv(ttr, relapse) ~ ageGroup4 + employment, data = dat)
```

## Comparing nested models: LRT
```{r}
anova(MA, MC)
```

## Comparing non-nested models: AIC
```{r}
fits <- list(A = MA, B = MB, C = MC)
sapply(fits, AIC)
```

```{r}
AIC(MC)
```


## Automatic model selection based on AIC

```{r}
Mfull <- coxph(Surv(ttr, relapse) ~ grp + gender + race +
                 employment + yearsSmoking + levelSmoking +
                 ageGroup4 + priorAttempts + longestNoSmoke,
               data = dat)
```

```{r}
MAIC <- step(Mfull)
```

```{r}
summary(MAIC)
```

```{r}
table(dat$ageGroup4)
```


## Predictive power: concordance index

```{r}
summary(MA)
```

```{r}
summary(MAIC)
```

## Predictive power: AUC

```{r}
library(survivalROC)
data(mayo)
```

```{r}
fit <- coxph(Surv(time, censor) ~ I(mayoscore5/sd(mayoscore5)), data = mayo)
summary(fit)
```


```{r}
plot(survfit(Surv(time / 365.25, censor) ~ 1, data = mayo))
```

```{r}
ROC.4 <- survivalROC(Stime = mayo$time,
                     status = mayo$censor,
                     marker = mayo$mayoscore4,
                     predict.time = 365.25 * 5,
                     method="KM")
ROC.5 <- survivalROC(Stime = mayo$time,
                     status = mayo$censor,
                     marker = mayo$mayoscore5,
                     predict.time = 365.25 * 5,
                     method = "KM")
```

```{r}
ROC <- list(mayo4 = ROC.4, mayo5 = ROC.5)
sapply(ROC, "[[", "AUC")
```

```{r}
dfl <- lapply(ROC, function(x) with(x, data.frame(FP, TP)))
for(nm in names(dfl)) {
  dfl[[ nm ]]$marker <- nm
}
dat <- do.call(rbind, dfl)
```

```{r}
library(ggplot2)
ggplot(dat, aes(FP, TP, color = marker)) +
  geom_line() +
  theme_bw(base_size = 9)
```

```{r}
cutoff <- with(ROC$mayo5, min(cut.values[FP <= 0.1]))
cutoff
```

```{r}
mayo$prediction <-
  ifelse(mayo$mayoscore5 <= cutoff,
         "low_risk", "high_risk")
```

```{r}
plot(survfit(Surv(time/365, censor) ~ prediction, data = mayo),
     col = c("red", "blue"))
```

# Model diagnostics
## Martingale residuals

```{r}
library(survival)
library(asaur) ## dataset

data(pharmacoSmoking)
dat <- pharmacoSmoking
```

```{r}
fit <- coxph(Surv(ttr, relapse) ~ grp + age + employment, data = dat)
dat$residual <- residuals(fit, type = "martingale")
```

```{r}
par(mfrow = c(1, 3), mar = c(4.2, 2, 2, 2))
with(dat, {

  plot(age, residual)
  lines(lowess(age, residual), lwd = 2)

  plot(residual ~ grp)

  plot(residual ~ employment)

})
```

```{r}
dfbetas <- residuals(fit, type = 'dfbetas')
dat$dfbetas <- sqrt(rowSums(dfbetas^2))
```

```{r}
plot(dat$dfbetas, type = 'h')
abline(h = 0)
```

## Proportionality of hazards

Pancreatic cancer dataset
```{r}
library(survival)
library(asaur) ## dataset
library(plyr)
library(ggplot2)

data(pancreatic)
dat <- pancreatic

fmt <- "%m/%d/%Y"
dat <- within(dat, {
  onstudy <- as.Date(as.character(onstudy), format = fmt)
  progression <- as.Date(as.character(progression), format = fmt)
  death <- as.Date(as.character(death), format = fmt)
  OS <- death - onstudy
  PFS <- pmin(progression - onstudy, OS)
  PFS[is.na(PFS)] <- OS[is.na(PFS)]
  PFS <- Surv(as.numeric(PFS / 30.5))
  OS <- Surv(as.numeric(OS / 30.5))
})
```

```{r}
fit <- coxph(PFS ~ stage, data = dat)
summary(fit)
```


```{r}
fit.KM <- survfit(PFS ~ stage, data = dat)
plot(fit.KM, fun= "cloglog", col = 1:2)
```


```{r}
fit.KM <- survfit(Surv(ttr, relapse) ~ grp, data = pharmacoSmoking)

plot(fit.KM, fun = "cloglog", col = 1:2)
```

### Schoenfeld residuals
```{r}
fit <- coxph(PFS ~ stage, data = dat)
residual.sch <- cox.zph(fit)
residual.sch
```


```{r}
plot(residual.sch)
```

# Dealing with assumptions violations

## Stratification

```{r}
library(asaur)
d <- pharmacoSmoking
d$employment <- ifelse(d$employment == "ft", "ft", "other")

table(d$employment)
```

Stratified Cox model:
```{r}
fit <- coxph(Surv(ttr, relapse) ~ grp + strata(employment), data = d)

summary(fit)
```

Note how there is no estimate associated with 'employment'.

## Truncation

```{r}
library(asaur)
library(survival)
d <- pancreatic2

plot(survfit(Surv(pfs, status) ~ stage, data = d), col = 1:2)
```

THIS IS *NOT* HOW IT IS DONE:
```{r}
d_WRONG <- subset(d, pfs <= 180)

plot(survfit(Surv(pfs, status) ~ stage, data = d_WRONG), col = 1:2)
```

Here is how you do it:
```{r}
d_RIGHT <- within(d, {
  status_truncated <- ifelse(pfs > 180, 0, status)
  pfs_truncated <- ifelse(pfs > 180, 180, pfs)
})
```

```{r}
plot(survfit(Surv(pfs_truncated, status_truncated) ~ stage, data = d_RIGHT),
     col = 1:2)
```

```{r}
plot(survfit(Surv(pfs, status) ~ stage, data = d_RIGHT),
     fun = "cloglog",
     col = 1:2)
```

```{r}
summary(coxph(Surv(pfs, status) ~ stage, data = d_RIGHT))
```

## Psplines

```{r}
library(asaur)
library(survival)
dat <- pharmacoSmoking

fit <- coxph(Surv(ttr, relapse) ~ grp + employment + age, data = dat)

summary(fit)
```

```{r}
fit1 <- coxph(Surv(ttr, relapse) ~ grp + employment + pspline(age, df = 4), data = dat)

summary(fit1)
```

```{r}
termplot(fit1, term = 3, se = TRUE)
```
