---
title:      "Survival Analysis project - Spring 18"
authors:    Mélanie 
            Pras
            Lucas
---
Steps
 - isolate 106 cases
 - residual NAs <- predict (lm)
 - exploratory/questions

Questions
 - treatment or not
 - global cox on all parameters
 - isolate significative parameters
 - generate model
 - predict NAs survival
 
Check
 - cholesterol vs. age???

# Data preparation

    Variable    Description
    ----------  -----------------------------------
    age:        in years
    albumin:	  serum albumin (g/dl)
    alk.phos:	  alkaline phosphotase (U/liter)
    ascites:	  presence of ascites
    ast:	      aspartate aminotransferase, once called SGOT (U/ml)
    bili:	      serum bilirunbin (mg/dl)
    chol:	      serum cholesterol (mg/dl)
    copper:	    urine copper (ug/day)
    edema:	    0 no edema, 0.5 untreated or successfully treated
                1 edema despite diuretic therapy
    hepato:	    presence of hepatomegaly or enlarged liver
    id:	        case number
    platelet:	  platelet count
    protime:	  standardised blood clotting time
    sex:	      m/f
    spiders:	  blood vessel malformations in the skin
    stage:	    histologic stage of disease (needs biopsy)
    status:	    status at endpoint, 0/1/2 for censored, transplant, dead
    time:	      number of days between registration and the earlier of death,
                transplantion, or study analysis in July, 1986
    trt:	      1/2/NA for D-penicillmain, placebo, not randomised
    trig:	      triglycerides (mg/dl)
    ----------  -----------------------------------

## Libraries importation
```{r}
library(ggplot2)
library(readr)
library(glmnet)
library(survival)
?pbc
```

## Data visualizations
```{r}
#head(pbc)
#summary(pbc)
#hist(pbc$stage)
#table(pbc$stage)
```

## Specify event as status == "death"
```{r}
data <- pbc
data$death <- 0 + (data$status == 2)
data$transp <- 0 + (data$status == 1)
#summary(data)
#table(data$status)
#table(data$event)
```

## Convert all categorical data of the data set
```{r}
data$trt <- factor(data$trt)
data$status <- factor(data$status)
data$stage <- factor(data$stage)
data$ascites <- factor(data$ascites)
data$edema <- factor(data$edema)
data$spiders <- factor(data$spiders)

#summary(data)
```

## Create age intervals
```{r}
data$ageGroup <- cut(data$age, breaks = c(0,10,20,30,40,50,60,70,80,90,Inf))

#summary(data)
```

## Isolate train data (312 observations) & test data (106 isolated cases that did not take any treatment)
```{r}
test <- subset(data, is.na(data$trt))
train <- subset(data, data$trt != "NA")

#summary(data)
#summary(test)
#summary(train)
```

## Treat lasting NAs in the train data set by predicting them linearly (vs. age)
```{r}
# cholesterol
fit.chol <- (lm(chol ~ age, data = train))
train$chol[is.na(train$chol)] <-
  predict(fit.chol, newdata = subset(train, is.na(chol)))
# copper
fit.copper <- (lm(copper ~ age, data = train))
train$copper[is.na(train$copper)] <-
  predict(fit.copper, newdata = subset(train, is.na(copper)))
# trig
fit.trig <- (lm(trig ~ age, data = train))
train$trig[is.na(train$trig)] <-
  predict(fit.trig, newdata = subset(train, is.na(trig)))
# platelet
fit.platelet <- (lm(platelet ~ age, data = train))
train$platelet[is.na(train$platelet)] <-
  predict(fit.platelet, newdata = subset(train, is.na(platelet)))

#summary(train)
```
By having separated the complete pbc data set into one train set of the 312 observations of patients that took a treatment (peniciillmain vs placebo) on one hand and a test set of 106 observations of patients that never took a treatment on the other hand, additionally to habing managed the residual NAs contained in the train test for full data cleansing we allow ourself to proceed to exploratory analysis.

# Exploratory analysis

## integrate a dependent variable y as the survival in months to data set
```{r}
train$y <- with(train, Surv(time / 30.5, death))
train$z <- with(train, Surv(time / 30.5, transp))
head(train)
summary(train$y)
summary(train$z)
```

## Create a survival objects
With 2 time ranges, in days as by default in the data set and in months as computed ad hoc
```{r}
survdays <- Surv(train$time, train$death)
survmonths <- Surv(train$time / 30.5, train$death)
#survdays
#survmonths
```

## Kaplan-Meyer estimator
Also known as product-limit estimator, like the censoring version of empirical survival function, generates a stair-step curve, but for effect of other covariates.
```{r}
KMdays <- survfit(survdays ~ 1, data = train)
KMmonths <- survfit(survmonths ~ 1, data = train)
plot(KMdays)
plot(KMmonths)
```

## Mantel-Haenzel Test
Also known as log-rank test, generated from a sequence of 2×2 tables measuring conditional independence.
Efficient in comparing groups differed by categorical variables, but not continuous ones.
```{r}
logrankdays <- survdiff(survdays ~ train$trt)
logrankmonths <- survdiff(survmonths ~ train$trt)
#plot(logrankdays)
#plot(logrankmonths)
```

## Cox Model
Also known as proportional hazard model, conveniently accesses the effect of continuous and categorical variable using partial likelihood to get inference even without knowledge of baseline hazard.
Note that the assumption is strong.
```{r}
fit.coxph.days <- coxph(survdays  ~ age + edema + log(bili) + log(albumin) + log(protime), data = train)
fit.coxph.months <- coxph(survmonths  ~ age + edema + log(bili) + log(albumin) + log(protime), data = train)
summary(fit.coxph.days)
summary(fit.coxph.months)
```

Plots
```{r}
data.null<-data.frame(age=rep(0,1), edema=rep(0,1), bili=rep(1,1), albumin=rep(1,1),protime=rep(1,1))

#plot(survfit(fit.coxph.days, newdata=data.null), lwd=2,ylim=c(.99,1), main='baseline survivor', xlab='Days', ylab='Survival', conf.int=T)
#plot(survfit(fit.coxph.months, newdata=data.null), lwd=2,ylim=c(.99,1), main='baseline survivor', xlab='Months', ylab='Survival', conf.int=T)

plot(survfit(fit.coxph.days),lwd=2,main= 'fitted survival function at mean covariates', xlab='Days', ylab='Survival')
plot(survfit(fit.coxph.months),lwd=2,main= 'fitted survival function at mean covariates', xlab='Months', ylab='Survival')
```

## Diagnostic of Cox Model
```{r}

```
