---
title:      "Survival Analysis project - Spring 18"
authors:    Melanie 
            Pras
            Lucas
---

Steps 
 - DONE // KM estimator 
 - Plot logrank curves
 - Plot cox baseline curves
 - check logrank curves pour troncate (juste un commentaire a faire), check jour Antonio Truncation video
 - chain les coxph pour identifier les variables significatives
 - analyse de residuals
 - bonus: comparaison de modeles
 
Questions
 - integrer les 19 patients transplantess dans les censures?
 - coxph au debut pour identifier les parametres significatifs?
 - interpretation MAPE & R²?
 - on veut predire le traitement ou la mort des 106?
 
-------------------------------------------------------------------------

# Introduction

Multiple tools and data for survival analysis are available in R packages such as "survival" from where we picked the PBC dataset. It comes from a clinical trial in the field of primary biliary cirrhosis conducted at the Mayo Clinic between 1974 and 1984. Primary biliary cirrhosis is a fatal chronic liver disease.

A total of 418 PBC patients were randomized to either a placebo or a drug called Dpenicillamine. Each of them was followed until death or censoring (the duration is measured in days). The status at endpoint is coded as follows: 0/1/2 for censored, transplant and dead respectively. In addition, 17 covariates are recorded for this study. These include a treatment variable, patient age, gender and clinical, biochemical and histologic measurements made at the time of randomization.

In this work, we will mainly consider the following variables: age (in years), serum albumin (g/dl), serum bilirubin (mg/dl), edema (0 if no edema, 0.5 if untreated or successfully treated and 1 if edema despite diuretic therapy) and prothrombin time (standardised blood clotting time).

# Data preparation

    Variable    Description
    ----------  -----------------------------------
    age:        in years
    albumin:	  serum albumin (g/dl)
    alk.phos:	  alkaline phosphotase (U/liter)
    ascites:	  presence of ascites
    ast:	      aspartate aminotransferase, once called SGOT (U/ml)
    bili:	      serum bilirunbin (mg/dl)
    chol:	      serum cholesterol (mg/dl)
    copper:	    urine copper (ug/day)
    edema:	    0 no edema, 0.5 untreated or successfully treated
                1 edema despite diuretic therapy
    hepato:	    presence of hepatomegaly or enlarged liver
    id:	        case number
    platelet:	  platelet count
    protime:	  standardised blood clotting time
    sex:	      m/f
    spiders:	  blood vessel malformations in the skin
    stage:	    histologic stage of disease (needs biopsy)
    status:	    status at endpoint, 0/1/2 for censored, transplant, dead
    time:	      number of days between registration and the earlier of death,
                transplantion, or study analysis in July, 1986
    trt:	      1/2/NA for D-penicillmain, placebo, not randomised
    trig:	      triglycerides (mg/dl)
    ----------  -----------------------------------

## Libraries importation

We also ground the present document on various other packages for data presentation and visualization (ie. gglopt2, readr, glmnet ...etc.).

```{r}
library(ggplot2)
library(survminer)
library(readr)
library(glmnet)
library(survival)
```

## Data visualizations

```{r}
#head(pbc)
#summary(pbc)
#hist(pbc$stage)
#table(pbc$stage)
```

## Specify event as status == "death"

Declare data importation and event association to the death of the patient.
Transplantation cases will not be of concerned in the context of a survival survival analysis, stricto sensu.

```{r}
# assign data set to a labelled object
data <- pbc
# create event parameter corresponding to death of the patient
data$event <- 0 + (data$status == 2)
```

## Convert the future covariates into factor

```{r}
data$trt <- factor(data$trt)
data$status <- factor(data$status)
data$stage <- factor(data$stage)
data$ascites <- factor(data$ascites)
data$edema <- factor(data$edema)
data$spiders <- factor(data$spiders)
```

## Create age intervals

Printing the histogram of the age variable might help naively identifying gaps or cuts in its distribution. This would highlight the number of modes encompassed in the variable. Obvious gaps would suggest cutoffs on which basing age groups variable.

```{r}
hist(data$age)
```

We observe a properly distributed age variable with no cuts or gaps thus implying a ageGroup variable evenly distributed too.

```{r}
data$ageGroup <- cut(data$age, breaks = c(0,10,20,30,40,50,60,70,80,90,Inf))
table(data$ageGroup)
```

## Input some missing values

Few missing values including for some future covariates invited us to apply linear model predictions.

```{r}
# cholesterol
fit.chol <- (lm(chol ~ age, data = data))
data$chol[is.na(data$chol)] <-
  predict(fit.chol, newdata = subset(data, is.na(chol)))
# copper
fit.copper <- (lm(copper ~ age, data = data))
data$copper[is.na(data$copper)] <-
  predict(fit.copper, newdata = subset(data, is.na(copper)))
# trig
fit.trig <- (lm(trig ~ age, data = data))
data$trig[is.na(data$trig)] <-
  predict(fit.trig, newdata = subset(data, is.na(trig)))
# platelet
fit.platelet <- (lm(platelet ~ age, data = data))
data$platelet[is.na(data$platelet)] <-
  predict(fit.platelet, newdata = subset(data, is.na(platelet)))
# check if NAs have been properly predicted
summary(data)
```

# Exploratory analysis

## Locate patients for survival analysis

Our patient's type-profile encompasses :
 - patients that followed a treatment, thus excluding the 106 patients that did not;
 - patients concerned by the event of their death or consored, thus excluding the patients that were transplanted.

```{r}
specimen <- subset(data, data$trt != "NA" & data$status != 1)
summary(specimen)
```

Getting a 293 observations data set on which we will run the survival analysis related tests as below.

## Create survival objects

Survival objects are created through the Surv(time, status) function from the "survival" package. To create right-censored data, this function needs two arguments:
 - time: returns the observed duration in days;
 - status: returns a boolean regarding whereas the observation corresponds to a censored one or not.

In the situation where status returns more than two modalities or if the modalities are not returning a boolean conditioned by the fact that the observations are censored or not, the formula creating the survival object must precise the proper modalities corresponding to censored observations.

```{r}
survival <- Surv(specimen$time / 365.25, specimen$event)
```

Hereby computed with time alteration to show yearly-basis scale for lisibility purpose of the reader.

## Kaplan-Meyer estimator - estimation of the survival function

Also known as "product-limit estimator", the Kaplan-Meyer estimator (KM) is a non-parametric statistic (ie. not based on the assumption of an underlying probability distribution) that allows us to estimate the survival function. It gives the probability that an individual patient will survive past a particular time t. It is based on the assumption that the probability of surviving past a certain time point t is equal to the product of the observed survival rates until time point t.
It is similar to the censoring version of empirical survival function, generating a stair-step curve but not accounting for effect of other covariates.
In R, the estimation of a survival function through the use of a survival object (ie. from censored data) is done thanks to the survfit(Surv(time, status), data) function of the "survival" package.

```{r}
KM <- survfit(survival ~ 1, data = specimen)
KM
ggsurvplot(KM, xlab="Survival Years", ylab="Kaplan-Meyer Estimator")
```

The KM test here returns a median survival of 8.99 years, the moment at which 50% of the patients were alive and 50% were reaching the event point ie. here, death. On a broader note, the reader may be interested in visualizing the survival regarding other parameters. This can be realised by crossing the survival object with the specific parameters through additional KM tests. Interesting parameters to be confronted to survival may be the sex parameter, the treatment parameter and the age parameter, the later requiring a preparation to its study (ie. "binarizing" the sample into "younger" vs "older" patients for example).

Considering the sex parameter first :

```{r}
# fitting the survival to sex parameter
fit.sex <- survfit(survival ~ sex, data = specimen)
## visualizing the survival probability
ggsurvplot(fit.sex, 
           data = specimen, 
           xlab = "Years",
           conf.int = FALSE,
           pval = TRUE,
           legend = "top",
           legend.title = "Sex",
           legend.labs = c("Male", "Female"))
## visualizing the cumulative event
ggsurvplot(fit.sex, 
           data = specimen, 
           xlab = "Years",
           fun = "event",
           legend = "top",
           legend.title = "Sex",
           legend.labs = c("Male", "Female"))
## visualizing the cumulative hazard
ggsurvplot(fit.sex, 
           data = specimen, 
           xlab = "Years",
           fun = "cumhaz",
           legend = "top",
           legend.title = "Sex",
           legend.labs = c("Male", "Female"))
```

Additionally to the general shape of the curves, the reader might be interested in the p-value shown at the bottom-left of the figure which is the corresponding log-rank test p-value result. Here statistically significant, as under an arbitrary threshold of 5% (95% confidence interval) we have enough significance to reject the log-rank null hypothesis and affirm that the two groups here male & female, survive differently to the biliary cirrhosis.

Now considering the treatment parameter (D-penicillmain or placebo) :

```{r}
# fitting the survival to treatment parameter
fit.trt <- survfit(survival ~ trt, data = specimen)
## visualizing the survival probability
ggsurvplot(fit.trt, 
           data = specimen, 
           xlab = "Years",
           conf.int = FALSE,
           pval = TRUE,
           legend = "top",
           legend.title = "Sex",
           legend.labs = c("Male", "Female"))
## visualizing the cumulative event
ggsurvplot(fit.trt, 
           data = specimen, 
           xlab = "Years",
           fun = "event",
           legend = "top",
           legend.title = "Sex",
           legend.labs = c("Male", "Female"))
## visualizing the cumulative hazard
ggsurvplot(fit.trt, 
           data = specimen, 
           xlab = "Years",
           fun = "cumhaz",
           legend = "top",
           legend.title = "Sex",
           legend.labs = c("Male", "Female"))
```

As visually as supported by the very high p-value resulting from the study, we do not have enough significance in order to reject the null hypothesis and thus conclude that there is no significant difference between both treatment protocols.

the provided p-value to the KM visualization of the survival object has introduced the reader to the observation of differences in some parameters variable to be explicited in the Mantel-Haenzel test, also called the log-rank test.

## Mantel-Haenzel test - comparing two groups' own survival

Also known as log-rank test, it is a statistical hypothesis test that tests the null hypothesis that survival curves of two populations do not differ. It will output an indicator of the two groups being significantly different in terms of survival when its p-value will be inferior to risk threshold.
It is generated from a sequence of 2 by 2 tables measuring conditional independence. It is efficient in comparing groups differed by categorical variables, but not continuous ones.

```{r}
MH <- survdiff(survival ~ specimen$trt)
MH
```

The log-rank returning a p-value way over the risk threshold c. 0.8 implies that we do not have enough elements to reject the null hypothesis as such there is no significant statistical difference between the two treatments.

Let's apply a Wilcoxon test in order to compare the significance result with the one from the log-rank test. The reader will be advised that however :
 - the log-rank test is more effective when the curves do not cross each other;
 - when instantaneous hazard rates are proportional, the log-rank test is the "best" to be run.
 
```{r}
W <- survdiff(survival ~ specimen$trt, rho=1)
W
```

Returning a p-value even higher than for the log-rank test preventing us from rejecting the null hypothesis one again and leading us to conclude that there is no statistically significant difference between the two studied groups D-penicillmain patients vs placebo ones.

## Cox Model

Also known as proportional hazard model, conveniently accesses the effect of continuous and categorical variable using partial likelihood to get inference even without knowledge of baseline hazard.
Note that the assumption is strong.

Whereas the log-rank test compares two Kaplan-Meier survival curves, which might be derived from splitting a patient population into treatment subgroups, Cox proportional hazards models are derived from the underlying baseline hazard functions of the patient populations in question and an arbitrary number of dichotomized covariates. Again, it does not assume an underlying probability distribution but it assumes that the hazards of the patient groups you compare are constant over time.

```{r}
# test ran with all covariates
coxph <- coxph(survival  ~ age + edema + hepato + platelet + sex + spiders + ascites + log(albumin) + log(alk.phos) + log(ast) + log(bili) + log(chol) + log(copper) + log(trig) + log(protime), data = specimen)
summary(coxph)
```

Various instructions can be used to adjust the Cox model on the data. Here above we adjusted the Cox model on our specimen data set with all the possible covariates the data set can possibly provide as predictors. The effects of each covariate are estimated. Our objective was to identify the covariates that are identified as significant predictors by the Cox model. We observe six covariates being significant to the data that we decide to select in order to refine our analysis.

```{r}
# fitted test with significant covariate only
fit.coxph <- coxph(survival  ~ age + as.factor(edema) + log(albumin) + log(bili) + log(protime) + log(copper), data = specimen)
summary(fit.coxph)
```

Another instruction could help us identify the interactions between the different covariates themselves and not only their unilateral effects of each one of them as predictors.

```{r}
# fitted test and interactions between covariates
fit.coxph.inter <- coxph(survival  ~ age * as.factor(edema) * log(albumin) * log(bili) * log(protime) * log(copper), data = specimen)
summary(fit.coxph.inter)
```

Visualizing the model.
The reader be advised that mentionning the as.factor() function on the edema parameter heped fix the plot of the baseline.

```{r}
specimen.null<-data.frame(age=rep(0,1), edema=rep(0,1), bili=rep(1,1), albumin=rep(1,1), protime=rep(1,1), copper=rep(1,1))
# for baseline
plot(survfit(fit.coxph, newdata=specimen.null), lwd=2,ylim=c(.99,1), main='Baseline survivor', xlab='Years', ylab='Survival', conf.int=T)
# for mean covariates
plot(survfit(fit.coxph),lwd=2,main= 'Fitted survivor at mean covariates', xlab='Years', ylab='Survival')
```

Here returning a unique curve for all the patients in the data set with a confidence interval.

Various other visualizations exist such as the function ggforest() from the survminer package which creates a forest plot for a Cox regression model fit. Hazard ratio estimates along with confidence intervals and p-values are plotter for each variable.

```{r}
ggforest(fit.coxph)
```

## Diagnostic of Cox Model

The function cox.zph() from survival package may be used to test the proportional hazards assumption for a Cox regression model fit. The graphical verification of this assumption may be performed with the function ggcoxzph() from the survminer package. For each covariate it produces plots with scaled Schoenfeld residuals against the time. 

```{r}
ftest <- cox.zph(fit.coxph)
ftest
ggcoxzph(ftest)
```

Additionally the function ggcoxdiagnostics() plots different types of residuals as a function of time, linear predictor or observation id. The type of residual is selected with type argument. Possible values are "martingale", "deviance", "score", "schoenfeld", "dfbeta"', "dfbetas", and "scaledsch". The ox.scale argument defines what shall be plotted on the OX axis. Possible values are "linear.predictions", "observation.id", "time". Logical arguments hline and sline may be used to add horizontal line or smooth line to the plot.

```{r}
ggcoxdiagnostics(fit.coxph,
                 type = "deviance",
                 ox.scale = "time")
ggcoxdiagnostics(fit.coxph,
                 type = "schoenfeld",
                 ox.scale = "time")
```

