---
title:      "Survival Analysis project - Spring 18"
authors:    Melanie 
            Pras
            Lucas
---
Steps
 - isolate 106 cases
 - residual NAs <- predict (lm)
 - exploratory/questions

Questions
 - treatment or not
 - global cox on all parameters
 - isolate significative parameters
 - generate model
 - predict NAs survival
 
Check
 - cholesterol vs. age???
 
Steps 
 - drop les predict actuels
 - split data set 312 : train(70%) validation(30%)
 - predict event death : test R² ou MAPE pour validation du modèle
 - predict sur les 106
 - check logrank curves pour trancate (juste un commentaire à faire), check jour Antonio Truncation video
 
-------------------------------------------------------------------------

# Introduction

Multiple tools and data for survival analysis are available in R packages such as "survival" from where we picked the PBC dataset. It comes from a clinical trial in the field of primary biliary cirrhosis conducted at the Mayo Clinic between 1974 and 1984. Primary biliary cirrhosis is a fatal chronic liver disease.

A total of 418 PBC patients were randomized to either a placebo or a drug called Dpenicillamine. Each of them was followed until death or censoring (the duration is measured in days). The status at endpoint is coded as follows: 0/1/2 for censored, transplant and dead respectively. In addition, 17 covariates are recorded for this study. These include a treatment variable, patient age, gender and clinical, biochemical and histologic measurements made at the time of randomization.

In this work, we will mainly consider the following variables: age (in years), serum albumin (g/dl), serum bilirubin (mg/dl), edema (0 if no edema, 0.5 if untreated or successfully treated and 1 if edema despite diuretic therapy) and prothrombin time (standardised blood clotting time).

# Data preparation

    Variable    Description
    ----------  -----------------------------------
    age:        in years
    albumin:	  serum albumin (g/dl)
    alk.phos:	  alkaline phosphotase (U/liter)
    ascites:	  presence of ascites
    ast:	      aspartate aminotransferase, once called SGOT (U/ml)
    bili:	      serum bilirunbin (mg/dl)
    chol:	      serum cholesterol (mg/dl)
    copper:	    urine copper (ug/day)
    edema:	    0 no edema, 0.5 untreated or successfully treated
                1 edema despite diuretic therapy
    hepato:	    presence of hepatomegaly or enlarged liver
    id:	        case number
    platelet:	  platelet count
    protime:	  standardised blood clotting time
    sex:	      m/f
    spiders:	  blood vessel malformations in the skin
    stage:	    histologic stage of disease (needs biopsy)
    status:	    status at endpoint, 0/1/2 for censored, transplant, dead
    time:	      number of days between registration and the earlier of death,
                transplantion, or study analysis in July, 1986
    trt:	      1/2/NA for D-penicillmain, placebo, not randomised
    trig:	      triglycerides (mg/dl)
    ----------  -----------------------------------

## Libraries importation

We also ground the present document on various other packages for data presentation and visualization (ie. gglopt2, readr, glmnet ...etc.).

```{r}
library(ggplot2)
library(readr)
library(glmnet)
library(survival)
?pbc
```

## Data visualizations

```{r}
#head(pbc)
#summary(pbc)
#hist(pbc$stage)
#table(pbc$stage)
```

## Specify event as status == "death"

Declare data importation and event association to death of the patient.

/*
  transplantation cases will not be of concerned in the context of a survival
  survival analysis, stricto sensu.
*/

```{r}
data <- pbc
data$death <- 0 + (data$status == 2)
data$transp <- 0 + (data$status == 1)
#summary(data)
#table(data$status)
#table(data$event)
```

## Convert all categorical data of the data set

```{r}
data$trt <- factor(data$trt)
data$status <- factor(data$status)
data$stage <- factor(data$stage)
data$ascites <- factor(data$ascites)
data$edema <- factor(data$edema)
data$spiders <- factor(data$spiders)

#summary(data)
```

## Create age intervals

```{r}
data$ageGroup <- cut(data$age, breaks = c(0,10,20,30,40,50,60,70,80,90,Inf))
hist(ageGroup)
#summary(data)
```

## Isolate train data (312 observations) & test data (106 isolated cases that did not take any treatment)

```{r}
test <- subset(data, is.na(data$trt))
train <- subset(data, data$trt != "NA")

#summary(data)
#summary(test)
#summary(train)
```

## Treat lasting NAs in the train data set by predicting them linearly (vs. age)

```{r}
# cholesterol
fit.chol <- (lm(chol ~ age, data = train))
train$chol[is.na(train$chol)] <-
  predict(fit.chol, newdata = subset(train, is.na(chol)))
# copper
fit.copper <- (lm(copper ~ age, data = train))
train$copper[is.na(train$copper)] <-
  predict(fit.copper, newdata = subset(train, is.na(copper)))
# trig
fit.trig <- (lm(trig ~ age, data = train))
train$trig[is.na(train$trig)] <-
  predict(fit.trig, newdata = subset(train, is.na(trig)))
# platelet
fit.platelet <- (lm(platelet ~ age, data = train))
train$platelet[is.na(train$platelet)] <-
  predict(fit.platelet, newdata = subset(train, is.na(platelet)))

#summary(train)
```

By having separated the complete pbc data set into one train set of the 312 observations of patients that took a treatment (peniciillmain vs placebo) on one hand and a test set of 106 observations of patients that never took a treatment on the other hand, additionally to having managed the residual NAs contained in the train test for full data cleansing we allow ourself to proceed to exploratory analysis.

# Exploratory analysis

## integrate a dependent variable y as the survival in months to data set

```{r}
train$y <- with(train, Surv(time / 30.5, death))
train$z <- with(train, Surv(time / 30.5, transp))
head(train)
summary(train$y)
summary(train$z)
```

## Create survival objects

Survival objects are created through the Surv(time, status) function from the "survival" package. To create right-censored data, this function needs two arguments:
 - time: returns the observed duration in days;
 - status: returns a boolean regarding whereas the observation corresponds to a censored one or not.

In the situation where status returns more than two modalities or if the modalities are not returning a boolean conditioned by the fact that the observations are censored or not, the formula creating the survival object must precise the proper modalities corresponding to censored observations.

/*
  Here with two time ranges, in days as by default in the data set and in
  months as computed ad hoc for lisibility
*/

```{r}
survdays <- Surv(train$time, train$death)
#survmonths <- Surv(train$time / 30.5, train$death)
#survdays
#survmonths
```

## Kaplan-Meyer estimator - estimation of the survival function

Also known as "product-limit estimator", the Kaplan-Meyer estimator (KM) is a non-parametric statistic that allows us to estimate the survival function. It gives the probability that an individual patient will survive past a particular time t. It is based on the assumption that the probability of surviving past a certain time point t is equal to the product of the observed survival rates until time point t.
It is similar to the censoring version of empirical survival function, generating a stair-step curve but not accounting for effect of other covariates.
In R, the estimation of a survival function through the use of a survival object (ie. from censored data) is done thanks to the survfit(Surv(time, status), data) function of the "survival" package.

```{r}
KMdays <- survfit(survdays ~ 1, data = train)
KMmonths <- survfit(survmonths ~ 1, data = train)
plot(KMdays)
plot(KMmonths)
```

## Mantel-Haenzel test - comparing two groups' own survival

Also known as log-rank test, it is a statistical hypothesis test that tests the null hypothesis that survival curves of two populations do not differ. It will output an indicator of the two groups veing significantly different in terms of survival when its p-value will be inferior to risk threshold.
It is generated from a sequence of 2 by 2 tables measuring conditional independence. It is efficient in comparing groups differed by categorical variables, but not continuous ones.
In R, 

```{r}
logrankdays <- survdiff(survdays ~ train$trt)
logrankmonths <- survdiff(survmonths ~ train$trt)
#plot(logrankdays)
#plot(logrankmonths)
```

## Cox Model

Also known as proportional hazard model, conveniently accesses the effect of continuous and categorical variable using partial likelihood to get inference even without knowledge of baseline hazard.
Note that the assumption is strong.

```{r}
fit.coxph.days <- coxph(survdays  ~ age + edema + log(bili) + log(albumin) + log(protime), data = train)
fit.coxph.months <- coxph(survmonths  ~ age + edema + log(bili) + log(albumin) + log(protime), data = train)
summary(fit.coxph.days)
summary(fit.coxph.months)
```

Visualizing the model:

```{r}
data.null<-data.frame(age=rep(0,1), edema=rep(0,1), bili=rep(1,1), albumin=rep(1,1),protime=rep(1,1))

#plot(survfit(fit.coxph.days, newdata=data.null), lwd=2,ylim=c(.99,1), main='baseline survivor', xlab='Days', ylab='Survival', conf.int=T)
#plot(survfit(fit.coxph.months, newdata=data.null), lwd=2,ylim=c(.99,1), main='baseline survivor', xlab='Months', ylab='Survival', conf.int=T)

plot(survfit(fit.coxph.days),lwd=2,main= 'fitted survival function at mean covariates', xlab='Days', ylab='Survival')
plot(survfit(fit.coxph.months),lwd=2,main= 'fitted survival function at mean covariates', xlab='Months', ylab='Survival')
```

## Diagnostic of Cox Model

```{r}

```
