---
title: "Case Study: duration of hospitalization"
---

# Data preparation

    Variable   Description
    ---------- -----------------------------------
    DUR        Duration of hospitalization (days)
    AGE        Age (years)
    SEX        1 = male 2 = female
    TEMP       Body temperature (degrees Fahrenheit)
    WBC        White blood cells per 100 ml blood
    ANTIB      Antibiotic use: 1 = yes 2 = no
    CULT       Blood culture taken 1 = yes 2 = no
    SERV       Service: 1 = medical 2 = surgical
    ---------- -----------------------------------

```{r}
library(ggplot2)
library(readr)
raw <-
  read_tsv("hospitalization.csv",
           col_types = cols(id = 'c',
                            dur = 'd',
                            age = 'd',
                            sex = 'c',
                            temp = 'd',
                            wbc = 'd',
                            antib = 'c',
                            cult = 'c',
                            serv = 'c'))

h <- within(raw, {
  sex <- factor(sex, levels = c('1', '2'), labels = c('m', 'f'))
  antib <- factor(antib, levels = c('2', '1'), labels = c('no', 'yes'))
  cult <- factor(cult, levels = c('2', '1'), labels = c('no', 'yes'))
  serv <- factor(serv, levels = c('1', '2'), labels = c('medical', 'surgical'))
  temp <- (temp - 32.0) * 5/9  ## convert to Celsius
})
h <- subset(h, select = -id)
```

# Exploratory analysis

```{r}
summary(h)
```

```{r}
hist(h$dur)
```

```{r}
Fn <- ecdf(h$dur)
summary(Fn)
```

```{r}
plot(Fn)
```

```{r}
hist(h$temp)
```

# Q1. How many patients go through overnight hospitalization?

```{r}
table(h$dur == 0)
```

```{r}
prop.table(table(h$dur == 0))
```

Overall, 12% of the patients did not go through overnight hospitalization.

We can get a confidence interval using the Binomial test:
```{r}
with(h, binom.test(table(dur == 0)))
```

# Q2. Is the body temperature at admission predictive of the duration of the hospitalization?

```{r}
plot(dur ~ temp, data = h)
```

Meh!

```{r}
ggplot(h, aes(x = temp > 37, y = dur)) +
  geom_boxplot()
```
The overlapping of the 2 boxes suggests that there is no additionnal arguments in favor ofthe fact that temperature would predict hospitalization duration.

```{r}
with(h, table(temp > 37, dur > 0, deparse.level = 2))
```
We only have 6 patients with high temperature & longer hospitalization duration.

```{r}
with(h, prop.table(table(temp > 37, dur > 0), 1))
```

# Q3. What about blood works?

```{r}
ggplot(h, aes(x = cult, y = dur)) +
  geom_boxplot()
```

```{r}
with(h, table(cult, dur > 0, deparse.level =2))
```

```{r}
with(h, prop.table(table(cult, dur > 0), 1))
```

# Q4. Statistical Modeling

## Data preparation
```{r}
h$y <- h$dur > 0
mean_temperature <- mean(h$temp)
h$x <- h$temp - mean_temperature
```

## The Likelihood function
```{r}
logLik <- function(theta) {
  alpha <- theta[1]
  beta <- theta[2]
  linear_predictor <- alpha + beta * h$x
  probabilities <- plogis(linear_predictor)
  log_terms <- dbinom(h$y,
                      size = 1,
                      prob = probabilities,
                      log = TRUE)
  sum(log_terms)
}
```

```{r}
debugonce(logLik)
logLik(c(3, 3))
```
debugonce will enter a step by step function code analysis the first time it is run, here on our function logLik.
The function will return a (unique) numerical answer, here -7.568636 for alpha = beta = 3.

## Numerical maximization
```{r}
library(maxLik)

fit <- maxLik(logLik, start = c(alpha = 0, beta = 0))
summary(fit)
```
maxLik provides rich output & tests providing t-value & p-value allowing to assess the rejection of H0 or not. 
Here p-value associated with Ho : "beta = 0" is > 0.05 then we cannot reject H0 which means beta is not significantly different from 0. ie we see beta at 3.6667, able to vary of about 4 (variance = 2.0751²).
Test on alpha H0 : "alpha = 0" is not that interesting.
It confirms our intuition, there is no significant relation between temperature & probability of overnight hosp.

## Logistic regression

```{r}
fit2 <- glm(y ~ x, data = h, family = 'binomial')
summary(fit2)
```
glm = Generalized Linear Model including linear, logistic, poisson, binomial regressions...etc.
family = 'binomial' => to process logistic regression. If y is not binary 1/0, it will return an error.
We get the same results approx. (algo is the same).

## Model-based predictions

In the probability scale:
```{r}
predict(fit2,
        newdata = data.frame(x = 38 - mean_temperature),
        type = "response")
```
We want to predict the probability of overnight hospitalization of a patient that has a body temperature of 38°C that we compare to the avg in order to get its deviation.
type = "response" : y-axis of the logit function (probability)

This is the so-called 'linear predictor':
```{r}
predict(fit2,
        newdata = data.frame(x = 38 - mean_temperature),
        type = "link")
```
alpha + beta * (38 - mean_temperature) = 6.87
type = "link" : x-axis of the logit function

# Exercises with simulated data

## Data generation
```{r}
set.seed(1234)
rate <- 0.2
y <- rexp(100, rate = rate)
```
lambda (rate) set at 0.2 albeit slides require 0.5

## Mean and median
```{r}
mean(y)
1/rate
```
rate being lambda, the theoretical counerpart of the mean(y) is 1/lambda (estimator of mean), results land pretty close.

```{r}
median(y)
log(2)/rate
```
rate being lambda, the theoretical counerpart of the median(y) is log(2)/lambda (estimator of median), results land pretty close.

## ECDF
```{r}
F <- ecdf(y)
plot(F)
curve(pexp(x, rate = rate), col = "red", add = TRUE)
```
Notes
rexp : random from exponential
pexp : CDF of the exponential
dexp : Density function of the exponential
qexp (pexp^-1) : quantile function from exponential

## Survival function
```{r}
S <- function(t) 1 - F(t)
curve(S(x), from = 0, to = 30)
curve(pexp(x, rate = rate, lower.tail = FALSE), col = "red", add = TRUE)
```
lower.taill = TRUE : gives the p to be lower or equal (CDF)
lower.tail = FALSE : gives 1-p to be lower or equal so the p to be higher (or equal) which is survival

## Density and hazard functions

```{r}
f <- density(y, from = 0)
curve(dexp(x, rate = rate), col = "red", from = 0, to = 25)
lines(f)
```

```{r}
h_empirical <- f$y / S(f$x)
plot(f$x, h_empirical, type = "l")
abline(h = rate, col = "red")
```

## Minus-log survival function

```{r}
clog_empirical <- -log(S(f$x))
plot(f$x, clog_empirical, type = "l")
abline(a = 0, b = rate, col = "red")
```
This kind of linear trend for minus-log survival suggests a constant hazard assumption might be appropriate for your data.

## Maximum Likelihood Estimation

### The log-likelihood function
```{r}
logLik <- function(logRate) {
  sum(dexp(y, rate = exp(logRate), log = TRUE))
}
```

```{r}
f <- Vectorize(logLik)
curve(f(x), from = -3, to = 0)
```
Vectorize(logLik) allow a function that initially takes one input at a time to take a whole vector instead ie. logLik.

### Numerical Maximization
```{r}
library(maxLik)
fit <- maxLik(f, start = c(logRate = 0))
summary(fit)
```

```{r}
rate_MLE_numerical <- unname(exp(fit$estimate))
rate_MLE_numerical
```

## Comparing different estimators

### Mean and median

```{r}
c(theoretical = 1/rate,
  empirical = mean(y),
  MLE_numerical = 1/rate_MLE_numerical)
```

```{r}
c(theoretical = log(2) / rate,
  empirical = median(y),
  MLE_numerical = log(2) / rate_MLE_numerical)
```

### Survival function

```{r}
Fn <- ecdf(y)
S_empirical <- function(t) 1.0 - Fn(t)
S_MLE_numerical <- function(t) pexp(t, rate = rate_MLE_numerical, lower.tail = FALSE)
S_theoretical <- function(t) pexp(t, rate = rate, lower.tail = FALSE)

curve(S_empirical, from = 0, to = 40)
curve(S_MLE_numerical, col = "blue", add = TRUE)
curve(S_theoretical, col = "red", add = TRUE)
```

### Hazard function

```{r}
h_theoretical <- Vectorize(function(t) rate)
h_MLE_numerical <- Vectorize(function(t) rate_MLE_numerical)

curve(h_theoretical, from = 0, to = 40, col = "red",
      xlab = "time", ylab = "hazard")
curve(h_MLE_numerical, col = "blue", add = TRUE)
```

# Fit the exponential distribution to the hospital duration data

## MLE
```{r}
logLik <- function(logRate) {
  sum(dexp(h$dur, rate = exp(logRate), log = TRUE))
}
fit <- maxLik(logLik, start = 0)
summary(fit)
```

## A model linking body temperature to risk

### Loglikelihood function

```{r}
logLik <- function(theta) {
  alpha <- theta[1]
  beta <- theta[2]
  terms <- dexp(h$dur,
                rate = exp(alpha + beta * (h$temp - 36)),
                log = TRUE)
  sum(terms)
}
```

### MLE
```{r}
fit <- maxLik(logLik, start = c(alpha = 0, beta = 0))
summary(fit)
```
Negative beta = the higher the temperature the lower the risk of exiting the hospital or the higher the probability of hosp. duration.
Plus beta p-value being c. 0.4, it has absolutely no significance (we cannot reject H0 stating "beta is null").

### Q. What's the probability of staying 5 days or more?

```{r}
curve(pexp(5.0,
           rate = exp(fit$estimate[1] + fit$estimate[2] * x),
           lower.tail = FALSE),
      from = 0, to = 2,
      xlab = "body temperature (C) - 36 (centered)",
      ylab = "P(T > 5 | temperature)")
```
How the probabilities over 5 days of more changes

### How good is the exponential distribution for our data?

We can only proceed with some approximate, qualitative assessments.

Lets split the body temperature variable into just 2 levels: 'low' (temp<=37) and 'high' (temp>37), and compare empirical and theoretical complementary-log-log survival functions.

```{r}
h_low <- subset(h, 36 <= temp & temp <= 37)
h_high <- subset(h, 37 < temp & temp <= 38)
```
we create 2 data subsets from temperature intervals.

```{r}
clog <- function(y) {
  function(t) {
    S <- 1.0 - ecdf(y)(t)
    -log(S)
  }
}
```

```{r}
curve(clog(h_low$dur)(x),
      from = 0, to = 30, col = "blue",
      xlab = "time (days)",
      ylab = "-log(S(t))")
abline(a = 0, b = exp(fit$estimate[1] + fit$estimate[2] * (36.5 - 36)),
       col = 'blue', lty = 2)

curve(clog(h_high$dur)(x),
      from = 0, to = 30, col = "red", add = TRUE)
abline(a = 0, b = exp(fit$estimate[1] + fit$estimate[2] * (37.5 - 36)),
       col = 'red', lty = 2)
```
blue : lower temperature people
red : higher temperature people
dashes : models
lines : observations

The model fits the data rather poorly.

Trying more flexible models (e.g., Weibull) is left as an exercise for the reader.

## Application

## Linear Regression
```{r}
m <- lm(dur ~temp, data = h)
summary(m)
```
alpha = intercept
beta = temp (body temperature)
Both non significant as per the p-values
Do we have a linear relation between hosp. duration & body temperature? NO
  The closer temp goes to 0°C, the closer the duration of hosp. will converge to -104.413 days.
  For every +1°C in temp, you'll stay 2.986 days more in hosp. duration

An adjustment might be:
```{r}
m <- lm(log(dur+1) ~ temp, data = h)
summary(m)
```
For each +1°C in temp, the logarithm of hosp. duration will increase by 0.5087 consisting in a positive trend although it doesn't make a lot of sense considering the tests

Visualizing it:
```{r}
plot(log(dur+1) ~temp, data = h)
abline(coef(m))
```

If we want to be more precise a multivariate linear regression model might be interesting:
```{r}
m <- lm(dur ~ temp + age, data = h)
summary(m)
```
MLR tells age is significant thus >< 0 as per its p-value, age of a patient starts to be a possibily interesting variable to explain hosp. duration.
The interecept of -219.54 is still non sensical thus the model doesn't extrapolate at all.

And comparing with the generalized linear model:
```{r}
m <- glm(I(dur > 0) ~ temp + age, data = h, family = 'binomial')
summary(m)
```
Here age is not quite significant contrary to temperature but put all together it kind of works a bit better. However wu should retain it as the temp-only model seen earlier doesn't explain the observation quite well.
The intercept it still doesn't extrapolate well: a new born (age == 0) with 0°C body temp would imply a hosp. duration of -271 days.

```{r}
plogis(-271)
```
Logit (sigmoid) function returning probabilty associated with parameters of the model.

## EXERCISE
Use model 2 (glm) to predict to predict P(y > 0 | Temp = 38, Age = 45) #mean_age <- mean(h$age)
```{r}
temp <- h$temp
age <- h$age
mean_temp <- mean(temp)
mean_age <- mean(age)
predict(m,
        newdata = data.frame(temp = 38 - mean_temp),
        type = "response")
```
Erreur?? Quelle est la syntaxe??

Essai avec MLE:
```{r}
logLik2 <- function(theta) {
  alpha <- theta[1]
  beta1 <- theta[2]
  beta2 <- theta[3]
  MLR_predictor <- alpha + beta1 * temp + beta2 * age
  probabilities <- plogis(MLR_predictor)
  log_terms <- dbinom(h$y,
                      size = 1,
                      prob = probabilities,
                      log = TRUE)
  sum(log_terms)
}
```

```{r}
fit_MLR <- maxLik(logLik2, start = c(alpha = 0, beta1 = 0, beta2 = 0))
summary(fit_MLR)
```